---
title: "covid_19"
author: "Sharon Lee"
date: "5/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(readr)
library(forecast)
library(fpp2)
library(TTR)
library(caret)
library(corrplot)
library(e1071)
options(warn=-1)
```

## R Markdown file for Covid-19 in China project
Utilizing the Johns Hopkins and ACAPs dataset to predict confirmed cases in China. From the Johns Hopkins dataset, I will be including the following variables- confirmed, deaths, recovered cases in  time series. From the acaps dataset, I will be including the following variables- country, category, measure type, targeted pop group and date implemented. I will be combining those variables listed together into a dataframe for machine learning modeling. 

EDA:
1. Johns Hopkins Dataset
#read in time series confirmed global data 
```{r johns hopkins}
setwd('C:/USFMSHI/HS630/hw/assign6/COVID-19/csse_covid_19_data/csse_covid_19_time_series/')
time_series <- read.csv("time_series_covid19_confirmed_global.csv", header = T)
```

#extract out china data
```{r}
time_china <- time_series[which(time_series$Country.Region == "China"), ]
#time_china
```

#take out country, lat and long
```{r}
time_china <- time_china[, c(-2, -3, -4)]  
time_china <- time_china[,-1]
#time_china
```

#first day: 1/22
```{r}
days <- c(1:97)
colnames(time_china) <- days

#sum up to get accumulated cases of confirmed
China_confirmed <- c()
for(i in colnames(time_china)){
  #print(sum(time_china[,i])) 
  China_confirmed[i] <- sum(time_china[,i])
}
China_confirmed
```

#creating final dataframe- China
```{r}
China <- as.data.frame(China_confirmed)
colnames(China) <- "Confirmed"
China
```

#read in time series death global data 
```{r }
setwd('C:/USFMSHI/HS630/hw/assign6/COVID-19/csse_covid_19_data/csse_covid_19_time_series/')
time_series2 <- read.csv("time_series_covid19_deaths_global.csv", header = T)
#time_series2
```

#extract out china data
```{r}
time_china2 <- time_series2[which(time_series2$Country.Region == "China"), ]
#take out country, lat and long
time_china2 <- time_china2[, c(-2, -3, -4)]  
time_china2 <- time_china2[,-1]
#time_china2
```

#get accumulated death cases of china
```{r}
colnames(time_china2) <- days
China_deaths <- c()
for(i in colnames(time_china2)){
  #print(sum(time_china[,i])) 
  China_deaths[i] <- sum(time_china2[,i])
}

China_deaths
```

#add to China dataframe
```{r}
China$Deaths <- China_deaths
China
```

#read in time series recovered global data 
```{r}
setwd('C:/USFMSHI/HS630/hw/assign6/COVID-19/csse_covid_19_data/csse_covid_19_time_series/')
time_series3 <- read.csv("time_series_covid19_recovered_global.csv", header = T)
#time_series3
```

#extract out china data
```{r}
time_china3 <- time_series3[which(time_series2$Country.Region == "China"), ]
#take out country, lat and long
time_china3 <- time_china3[, c(-2, -3, -4)]  
time_china3 <- time_china3[,-1]

colnames(time_china3) <- days
China_recovered <- c()
for(i in colnames(time_china3)){
  China_recovered[i] <- sum(time_china3[,i])
}

China_recovered
China$Recovered <- China_recovered
China
```

#create columns that hold confirmed, death and recovered cases for each day instead of accumulation
```{r}
#confirmed
day_confirmed <- c()
for(i in (2:length(China$Confirmed))){
  
  day_confirmed[i] <- China$Confirmed[i]- China$Confirmed[i-1]
}
day_confirmed[1] <- China$Confirmed[1]
China$day_confirmed <- day_confirmed

#death
day_death <- c()
for(i in (2:length(China$Deaths))){
  
  day_death[i] <- China$Deaths[i]- China$Deaths[i-1]
}
day_death[1] <- China$Deaths[1]
China$day_death <- day_death

#recovered
day_rec <- c()
for(i in (2:length(China$Recovered))){
  
  day_rec[i] <- China$Recovered[i]- China$Recovered[i-1]
}
day_rec[1] <- China$Recovered[1]
China$day_recovered <- day_rec
China
```

#add day column
```{r}
startDate <- as.Date("2020-01-22")
endDate <- as.Date("2020-04-27")
date <- seq(startDate, endDate, "1 day")  
China$day <- date
```

2. ACAPS Dataset
```{r ACAPS}
setwd('C:/USFMSHI/HS630/hw/assign6/other_datasets')
safety <- read.csv("acaps.csv", header = T, stringsAsFactors = FALSE)
#safety
```

#keep variables that are necessary
```{r}
safety <- data.frame(safety$COUNTRY, safety$CATEGORY, safety$MEASURE, safety$TARGETED_POP_GROUP, safety$DATE_IMPLEMENTED, safety$SOURCE_TYPE)
#safety
```

#make a dataframe that only has data of china
```{r}
safety_china <- safety[which(safety$safety.COUNTRY == "China"), ]
safety_china
names <- c("Country", "Category", "Measure", "Targeted_by_pop", "Date_implemented", "Source_type")
colnames(safety_china) <- names
safety_china <- safety_china[,-1]
```

#data cleaning:
```{r}
summary(safety_china)
```

#things to be cleaned:Category, targeted_pop_group, date_implemented, source_type
#Category: get rid of the blank level
```{r}
lev_c <- c("Governance and socio-economic measures", "Humanitarian exemption", "Lockdown", "Movement restrictions", 
           "Public health measures", "Social distancing")
safety_china$Categories <- factor(safety_china$Category, ordered = F, levels = lev_c)
summary(safety_china$Categories)
safety_china <- safety_china[,-1]
```

#targeted_pop_group
```{r}
lev <- c("No", "Yes")
#class(safety_china$Targeted_by_pop)
safety_china$target_pop <- factor(safety_china$Targeted_by_pop, ordered = F, levels = lev)
safety_china <- safety_china[,-2]
```

#source type
```{r}
lev_s <- c("Government", "Media", "Other", "Other organisations", "Social media", "UN")
safety_china$Sources <- factor(safety_china$Source_type, ordered = F, levels = lev_s)
summary(safety_china$Sources)     #we have 3 NAs for this column
safety_china <- safety_china[,-3]
```

#date implemented
```{r}

summary(safety_china$Date_implemented)
#change date to date variable
safety_china$Date_implemented <- as.Date(safety_china$Date_implemented, format = "%d/%m/%Y")
safety_china$Date_implemented
safety_china <- safety_china[order(safety_china$Date_implemented),]
safety_china$Date_implemented
```

#create a dataframe that only contains the dates that we want (match China dataset)
```{r}
safety_china2 <- with(safety_china, safety_china[(Date_implemented >= "2020-01-22" & Date_implemented <= "2020-04-27"), ])
#remove NAs
safety_china2 <- na.omit(safety_china2)
```

#goal is to clean up the columns in order to combine with China dataframe. 
#issue now is: some dates have multiple entries of safety measures, and some dates have nothing at all
```{r}
#measures are too specific, take out measure col
safety_china2 <- safety_china2[,-1]

#renumbering rows
row.names(safety_china2) <- 1:nrow(safety_china2)
safety_china2$date <- c()
```

#start the process of adding safety categoy and target by pop columns to China dataframe
#added a never level for the category- no action taken for dates with no entry 
```{r}
cat_lev <- c("No action taken", "Social distancing", "Movement restrictions", "Public health measures", 
             "Governance and socio-economic measures", "Lockdown")

China$categories <- factor(rep(NA, nrow(China)), order = F, levels = cat_lev)
China$target <- factor(rep(NA, nrow(China)), order = F, levels = lev)
```

#combining safety category and date implemented to China dataframe
```{r}
startD = as.Date("2020-01-22")
endD = as.Date("2020-04-27")
theDate <- startD
while(theDate <= endD){
  #if the date is not in safety_china2, aka no safety action was implemented that day
  if(!(theDate%in%safety_china2$Date_implemented)){
    China$categories[China$day == theDate] <- "No action taken"
    China$target[China$day == theDate] <- "No"
  }
  #if the date is in date_implemented
  else if(theDate%in%safety_china2$Date_implemented){
    #only one entry 
    if(nrow(safety_china2[safety_china2$Date_implemented == theDate, ]) == 1){
      China$categories[China$day == theDate] <- safety_china2$Categories[safety_china2$Date_implemented == theDate]
      China$target[China$day == theDate] <- safety_china2$target_pop[safety_china2$Date_implemented == theDate]
    }
    #dates with multiple safety categories
    else{
      #maybe implement this if case later
      China$categories[China$day == theDate] <- safety_china2$Categories[safety_china2$Date_implemented == theDate]
      China$target[China$day == theDate] <- safety_china2$target_pop[safety_china2$Date_implemented == theDate]
    }
  }
  theDate = theDate + 1 
}
China
```
Dataset construction is complete. 

Data Visualization:
```{r}
g1 <- ggplot(China, aes(day, Confirmed))
g1 + geom_point(aes(col = day)) + 
  labs(title = "China confirmed cases over days")
```

```{r}
g2 <- ggplot(China, aes(day, Deaths))
g2 + geom_col(fill="#69b3a2") + 
  labs(title = "China death cases over days")
```

```{r}
g3 <- ggplot(China, aes(day,Recovered))
g3 + geom_line(col = "purple", size = 2) + 
  labs(title = "China recovered cases over days")
```


```{r}
g4 <-ggplot(China, aes(day, day_confirmed)) 
g4 + geom_line(col = "pink", size = 2) +
  labs(title = "Daily confirmed cases in China")
```

```{r}
g5 <- ggplot(China, aes(day, day_death))
g5 + geom_col(col = "orange", size = 0.5) +
  labs(title = "Daily death cases in China")
```

```{r}
g6 <- ggplot(China, aes(day, day_recovered))
g6 + geom_line(col = "blue", size = 1) + 
  labs(title = "Daily recovered cases in China")   ##there's a negative recovered day 

#trace back see how that happened
China[which(China$day_recovered < 0), ]
#switch these rows to NAs because the recovered cases spiked up in these days, then got adjusted 
#back down 
China[84:86, c(3,6)] <- NA
#re-adjust daily recovered cases for day 87
China[87, 6] <- China[87, 3] - China[83, 3]
China[83:87,]

```
These three graphs are the daily trends for confirmed, death and recovered cases in China. These graphs revealed a lot of interesting information about the dataset: first, as you can see, for the daily confirmed cases graph, you see a sudden big spike, and same thing for daily death cases- China went on days without any death cases reported, then all of a sudden, we see a huge spike of over 1500+ cases reported in one day. Lastly, for the recovered cases, there is even a negative spike of -600 seen here. Overall, these three graphs have shown me that there is a possibility of inconsistencies when it comes to reporting cases from China. At the end, I had to convert the entry of recovered and daily recovered to NAs for the 3 rows above the negative case in order to balance the data.

```{r}
g7 <- ggplot(China, aes(day_confirmed, day_recovered))
g7 + geom_smooth(col = "darkblue")
```

#plot some double plots
```{r}
g8 <- ggplot() +
  
  geom_col(data = China, aes(x = day, y = Confirmed), fill = "orange") + 
  geom_line(data = China, aes(x = day, y = Deaths),
            colour = "red", size = 2) +
  geom_point(data = China, aes(x = day, y=Recovered)) +
  labs(title = "China overall confirmed, death and recovered cases")

plot(g8)
```
overlapped confirmed, recovered and death cases to get a better idea about the relationship. The orange columns are the confirmed cases, blue dotted line is the recovered cases and red solid line represents death cases. 

#category
```{r}
g9 <- ggplot(data = China, aes(categories))
g9 + geom_bar(aes(fill=target), width = 0.5) + 
  theme(axis.text.x = element_text(angle=75, vjust=0.6)) + 
  labs(title="Histogram on safety categories in China") 
```

#targeted_pop_group
```{r}
g10 <- ggplot(China, aes(target))
g10 + geom_bar(aes(fill=categories), width = 0.75) + 
  theme(axis.text.x = element_text(angle=75, vjust=0.6)) + 
  labs(title="Histogram on targeted by population in China") 
```

```{r}
g11 <- ggplot(China, aes(categories,day)) 
g11 + geom_boxplot(aes(fill = target))+ 
  theme(axis.text.x = element_text(angle=75, vjust=0.6))+
  labs(title = "Box plot for safety cateogies and days")
```

```{r}
plot(China)
```

#correlation plot
```{r}
china_numeric <- China[, -c(7, 8, 9)]
china_numeric$categories <- as.numeric(China$categories)
china_numeric$target <- as.numeric(China$target)

c <- cor(china_numeric, use = "pairwise.complete.obs", method = "spearman")
c
corrplot(c)
```
Data Modeling: trying 3 dataframes with multiple modeling methods including linear regression, rpart, knn, svmRadial and random forest

1. China
```{r}
china_na <- na.omit(China)
is.na(china_na)

set.seed(3033)
intrain <- createDataPartition(y = china_na$Confirmed, p= 0.7, list = FALSE)
training <- china_na[intrain,]
testing <- china_na[-intrain,]
dim(intrain); dim(training); dim(testing)
```

#linear regression
```{r}

fit_all <- lm(Confirmed~ ., data = training)
summary(fit_all)
fit_null <- lm(Confirmed ~ 1, data = training)

fit_step <- step(fit_null, scope=list(lower=fit_null, upper=fit_all),direction="both")
summary(fit_step)
#adjusted R-squred of 0.89, really high 

plot(fit_step, 1) #very bad, seems to show a trend

p <- predict(fit_step, testing, type = "response")
p
plot(p)
```

#try different models to find the best one 
```{r}
trctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 3)
grid <- expand.grid(C = seq(0, 2, length =20))
```

#RPART
```{r}
set.seed(3233)
fit.cart <- train(Confirmed~., data = training, method = "rpart", 
                  preProcess = c("center", "scale"),
                  trControl = trctrl)
fit.cart
plot(fit.cart)
```

# kNN
```{r}
set.seed(3233)
fit.knn <- train(Confirmed~., data = training, method = "knn", 
                 preProcess = c("center", "scale"),
                 trControl = trctrl, 
                 tuneLength = 10)
fit.knn
plot(fit.knn)
#performance slightly better than rpart, but not that much better 
```

#SVMRadial
```{r}
trctrl_rbf <- trainControl(method = "repeatedcv", number = 5, repeats = 3)
svmGrid <- expand.grid(sigma= 2^c(-15,-10, -5, 0), C= 2^c(0:5))

set.seed(3233)

svm_Radial <- train(Confirmed ~., data = training, method = "svmRadial",
                    trControl=trctrl_rbf,
                    preProcess = c("center", "scale"),
                    tuneGrid = svmGrid,
                    tuneLength = 10)

svm_Radial   #extrememly high RMSE and MAE
plot(svm_Radial)

test_pred <- predict(svm_Radial, newdata = testing)
test_pred
plot(test_pred)
```

#Random Forest
```{r}
#using random search 
trctrl_rf <- trainControl(method="repeatedcv", number=10, repeats=3, search="random")
set.seed(3233)
mtry <- sqrt(ncol(training))
tunegrid <- expand.grid(.mtry=mtry)
rf_gridsearch <- train(Confirmed ~ ., data = training, method="rf", 
                       trControl=trctrl_rf,
                       preProcess = c("center", "scale"), 
                       tuneLength = 15)
print(rf_gridsearch) #best result so far  1518.527  0.9979277   873.7642
plot(rf_gridsearch)
```

```{r}
test_pred <- predict(rf_gridsearch, newdata = testing)
test_pred
plot(test_pred)
```

#For China dataset, random forest using random search produced the best result. 
#Use this method to predict the entire dataset:
```{r}
china_pred <- predict(rf_gridsearch, newdata = China)
china_pred
plot(china_pred, main = "Predicted Confirmed Cases in China Using Random Forest", xlab = "Days", 
     ylab = "Predicted Cases", col = "red")
```

2. China with Confirmed, Deaths, Recovered, dates, Category and Target
```{r}
China_2 <- China[, -c(4, 5, 6)]
China_2
China_2 <- na.omit(China_2)

set.seed(3033)
intrain <- createDataPartition(y = China_2$Confirmed, p= 0.7, list = FALSE)
training <- China_2[intrain,]
testing <- China_2[-intrain,]
dim(intrain); dim(training); dim(testing)
```

#linear regression
```{r}
fit_all <- lm(Confirmed~ ., data = training)
summary(fit_all)

fit_null <- lm(Confirmed ~ 1, data = training)

fit_step <- step(fit_null, scope=list(lower=fit_null, upper=fit_all),direction="both")
summary(fit_step)
plot(fit_step, 1) #horrible trend but high R-squared

p <- predict(fit_step, testing, type = "response")
p
plot(p)
```

#try different models to find the best one 
```{r}
trctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 3)
grid <- expand.grid(C = seq(0, 2, length =20))
```

#RPART
```{r}

set.seed(3233)
fit.cart <- train(Confirmed ~., data = training, method = "rpart", 
                  preProcess = c("center", "scale"),
                  trControl = trctrl)
fit.cart
plot(fit.cart)
```

#kNN
```{r}
set.seed(3233)
fit.knn <- train(Confirmed~., data = training, method = "knn", 
                 preProcess = c("center", "scale"),
                 trControl = trctrl, 
                 tuneLength = 10)
fit.knn
plot(fit.knn)
# k   RMSE       Rsquared   MAE     
#5   2829.361  0.9906999  1618.563
```

#SVMRadial
```{r}
trctrl_rbf <- trainControl(method = "repeatedcv", number = 5, repeats = 3)
svmGrid <- expand.grid(sigma= 2^c(-15,-10, -5, 0), C= 2^c(0:5))
set.seed(3233)
svm_Radial <- train(Confirmed ~., data = training, method = "svmRadial",
                    trControl=trctrl_rbf,
                    preProcess = c("center", "scale"),
                    tuneGrid = svmGrid,
                    tuneLength = 10)

svm_Radial   #rbf kernel is not the method for this model
plot(svm_Radial)

test_pred <- predict(svm_Radial, newdata = testing)
test_pred
```

#Random Forest
```{r}
#using random search 
trctrl_rf <- trainControl(method="repeatedcv", number=10, repeats=3, search="random")
set.seed(3233)
mtry <- sqrt(ncol(training))
tunegrid <- expand.grid(.mtry=mtry)
rf_gridsearch <- train(Confirmed ~ ., data = training, method="rf", 
                       trControl=trctrl_rf,
                       preProcess = c("center", "scale"), 
                       tuneLength = 15)
print(rf_gridsearch) #best result so far  1361.602  0.9980964    780.5278 (better than using entire China)
plot(rf_gridsearch)
```

#standardizes test data the same way as the training data 
```{r}
test_pred <- predict(rf_gridsearch, newdata = testing)
test_pred
```

#predict on entire dataset 
```{r}
test <- predict(rf_gridsearch, newdata = China)
test
plot(test, main = "Predicted Confirmed Cases in China Using Random Forest #2", xlab = "Days", 
     ylab = "Predicted Cases", col = "orange")
```

#plot predicted against actual confirmed cases
```{r}
china_predict <- data.frame(test)
#in order to overlay the predicted plot with the actual confirmed plot, need to fix the length (3 NA rows were omitted)
temp1 <- china_predict[1:83, ]
temp2 <- china_predict[84: nrow(china_predict), ]
temp3 <- rep(NA, 3)

china_pred <- data.frame(c(temp1, temp3, temp2))

colnames(china_pred) <- "Predicted_confirmed"
china_pred$date <- date
china_pred

f <- ggplot()+
  geom_line(data = China, mapping =aes(x = date, y = Confirmed), col = "orange", size = 1) + 
  geom_point(china_pred, mapping =aes(date, Predicted_confirmed), col = "blue")+
  labs(title = "Predicted (blue dots) and Actual (orange line) Confirmed Cases in China")

plot(f)
```

3. China with confirmed converted to log to see if prediction can be further improved
#china log
```{r}
China_log <- China 
China_log <- na.omit(China_log)
China_log$Confirmed <- log(China_log$Confirmed + 1)

set.seed(3323)
intrain <- createDataPartition(y = China_log$Confirmed, p= 0.7, list = FALSE)
training <- China_log[intrain,]
testing <- China_log[-intrain,]
dim(intrain); dim(training); dim(testing)
```

#rf
```{r}
set.seed(3233)
tc <- trainControl("oob") ##out of bag
Grid <- expand.grid(mtry = c(1:8))
fit.rf <- train(Confirmed ~., data = training , method = 'rf', trControl = tc, tuneGrid = Grid)
fit.rf  #best result: RMSE  0.24
```

```{r}
test_pred <- predict(fit.rf, testing)
test_pred <- exp(test_pred)-1
test_pred
plot(test_pred)
```

#predict against the whole dataset
```{r}
test <- predict(fit.rf, China)
test <- exp(test)-1
test
plot(test, main = "Predicted Confirmed Cases in China #3", xlab = "days", ylab = "Confirmed Cases", col = "blue")
```

#plot this predicted model against actual confirmed cases
```{r}
china_predict2 <- data.frame(test)

#in order to overlay the predicted plot with the actual confirmed plot, need to fix the length (3 NA rows were omitted)
temp1 <- china_predict2[1:83, ]
temp2 <- china_predict2[84: nrow(china_predict2), ]
temp3 <- rep(NA, 3)

china_pred2 <- data.frame(c(temp1, temp3, temp2))

colnames(china_pred2) <- "Predicted_confirmed"
china_pred2$date <- date
china_pred2

f2 <- ggplot()+
  geom_line(data = China, mapping =aes(x = date, y = Confirmed), col = "red", size = 1) + 
  geom_point(china_pred2, mapping =aes(date, Predicted_confirmed), col = "blue")+
  labs(title = "Predicted (blue dots) and Actual (orange line) Confirmed Cases in China #2")

plot(f2)
```

#plotting 2 predicted models with actual confirmed cases
```{r}
f3 <- ggplot()+
  geom_line(data = China, mapping =aes(x = date, y = Confirmed), col = "red", size = 1) + 
  geom_point(china_pred, mapping = aes(x = date, y = Predicted_confirmed), col = "green" ) + 
  geom_point(china_pred2, mapping =aes(date, Predicted_confirmed), col = "blue")+
  labs(title = "Predicted (green and blue) and Actual (red) Confirmed Cases in China ")
plot(f3)
```

Overall, regardless which dataframe used, random forest produced the best outcome. Random forest is able to give us the trend that closely matches the actual China confirmed cases data. When evaluating all the models, RMSE and MAE values were evaluated, and random forest for all cases produced the lowest RMSE and MAE by a huge margin. 
